---
title: "Project1"
author: "Ajay Tripathi"
date: "April 28, 2018"
output: html_document
---

## Loading and preprocessing the data

```{r pa1, echo=TRUE}
activity<-read.csv("./activity.csv",header = TRUE)
activity[,2]<-as.Date(activity$date)
```
## Now let's see how data looks like
```{r pa2, echo=TRUE}
str(activity)
```
## What is mean total number of steps taken per day?

Let's first build and see the histogram of total steps per day

```{r pa3, echo=TRUE}
steps_1<-with(activity,tapply(steps,date,sum,na.rm=TRUE))
hist(steps_1,col = "green",xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")
```




Now, let's see the average total steps and median of total steps taken each day and summary of the total steps taken each day

```{r pa4, echo=TRUE}
print(mean_steps<-mean(steps_1))
print(median_steps<-median(steps_1))
summary(steps_1)
```
## What is the average daily activity pattern?

Let's first make a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all days (y-axis).

```{r=pa5, echo=TRUE}
avg_steps<-with(activity,tapply(steps,interval,mean,na.rm=TRUE))
intervals<-unique(activity$interval)
new<-data.frame(cbind(avg_steps,intervals))
plot(new$intervals,new$avg_steps,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Average Steps per Interval")
```




Now, let's see which 5-minute interval contains the maximum number of steps
```{r=pa6,echo=TRUE}
index<-which.max(new$avg_steps)
max<-new[index,2]
```
##Imputing missing values
Calculate the average of average steps per day across all dates in the data set (ignoring NA values). Then use the resulting value in place of NAs.
Therefore, let's create a new dataset that is equal to the original dataset but with the missing data filled in.

```{r=pa7, echo=TRUE}
sum(is.na(activity$steps))
```
```{r=pa8,echo=TRUE}
index<-which(is.na(activity$steps))
l<-length(index)
steps_avg<-with(activity,tapply(steps,date,mean,na.rm=TRUE))
na<-mean(steps_avg,na.rm = TRUE)
for (i in 1:l) {
        activity[index[i],1]<-na
}
```



Let's see if we filled all NAs properly and see how our new dataset looks like
```{r=pa9,echo=TRUE}
sum(is.na(activity$steps))
str(activity)
```
## Let's see the histogram of total steps taken each day with the new dataset

```{r=pa10, echo=TRUE}
steps_2<-with(activity,tapply(steps,date,sum,na.rm=TRUE))
hist(steps_2,col = "green",xlab = "Total Steps",ylab = "Frequency",main = "Total Number of Steps per Day")
```



Let's see the mean and median of the total steps taken each day
```{r=pa11,echo=TRUE}
print(mean_steps_2<-mean(steps_2))
print(median_steps_2<-median(steps_2))
```
##Are there differences in activity patterns between weekdays and weekends?
In this section, we will use dplyr package and we need to load it from the library.
```{r=pa12, echo=TRUE}
library(dplyr)
activity_mod<- mutate(activity, day = ifelse(weekdays(activity$date) == "Saturday" | weekdays(activity$date) == "Sunday", "weekend", "weekday"))
activity_mod$day<-as.factor(activity_mod$day)
str(activity_mod)
act_wknd<-subset(activity_mod,as.character(activity_mod$day)=="weekend")
act_wkdy<-subset(activity_mod,as.character(activity_mod$day)=="weekday")
steps_wknd<-with(act_wknd,tapply(steps,interval,mean,na.rm=TRUE))
steps_wkdy<-with(act_wkdy,tapply(steps,interval,mean,na.rm=TRUE))
int_wknd<-unique(act_wknd$interval)
int_wkdy<-unique(act_wkdy$interval)
new_wknd<-data.frame(cbind(steps_wknd,int_wknd))
new_wkdy<-data.frame(cbind(steps_wkdy,int_wkdy))
par(mfrow=c(2,1),mar=c(4,4,2,1))
plot(new_wknd$int_wknd,new_wknd$steps_wknd,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Weekend")
plot(new_wkdy$int_wkdy,new_wkdy$steps_wkdy,type = "l",xlab = "Intervals",
     ylab = "Average Steps",main = "Weekday")
```




It is clear that the average steps over the weekends show higher pattern than that of the weekdays